---
title: "nfl_sports_reference_tutorial"
author: "Siddharth Venkiteswaran"
date: '2022-07-28'
output: html_document
---

## Step 1
Load in the libraries needed to extract and clean the data
```{r}
library(tidyverse)
library(readxl)
library(htmltab)
library(rvest)
```

## Step 2
Go into pro-football-reference.com and select the table you want to work with, and then import that data into R and turn it into a data frame

In this example, I will be using the data that contains the statistics of NFL Quarterbacks from the 2021 NFL season

Make sure you have the selector gadget Chrome extension to select the table you want to use
```{r}
## Create a URL for the passing data
passing_data_link <- read_html('https://www.pro-football-reference.com/years/2021/passing.htm')

## Turn the information in the URL into a list
passing_data_list = passing_data_link %>%
  ## You can find the label of the specific table by using the selector gadget to select the table you want to use
  html_nodes('#passing') %>%
  html_table()

## Create a temporary data frame to hold the information from the list
## This is a temporary data frame because we still have to clean the data once we bring the data in
temp_passing_df = passing_data_list[[1]]
```
