---
title: "nfl_sports_reference_tutorial"
author: "Siddharth Venkiteswaran"
date: '2022-07-28'
output: html_document
---
## Making Visualizations

## Step 1
Load in the libraries needed to extract and clean the data
```{r}
library(tidyverse)
library(readxl)
library(htmltab)
library(rvest)
library(bslib)
library(ggtext)
library(ggalt)
library(kableExtra)
```

```{r}
theme = bs_theme(
    bg = "#003831", 
    fg = "#FFE395", 
    primary = "#B38F4F", 
    base_font = font_google("Prompt"),
    code_font = font_google("JetBrains Mono")
  )
```

## Step 2
Go into pro-football-reference.com and select the table you want to work with, and then import that data into R and turn it into a data frame

In this example, I will be using the data that contains the statistics of NFL Quarterbacks from the 2021 NFL season

Make sure you have the selector gadget Chrome extension to select the table you want to use
```{r}
## Create a URL for the passing data
passing_data_link <- read_html('https://www.pro-football-reference.com/years/2021/passing.htm')

## Turn the information in the URL into a list
passing_data_list = passing_data_link %>%
  ## You can find the label of the specific table by using the selector gadget to select the table you want to use
  html_nodes('#passing') %>%
  html_table()

## Create a temporary data frame to hold the information from the list
## This is a temporary data frame because we still have to clean the data once we bring the data in
temp_passing_df = passing_data_list[[1]]
```

## Step 3
Once the data from the URL is turned into a data frame, it's time to clean the data so that we can use it to make visualizations and perform statistical methods with it
```{r}
## Remove the duplicate names in the temporary data frame
## We cannot clean the data until all of the columns have unique names
colnames(temp_passing_df)[c(11,14,16,17,19,20,21,22,26,27,28,29,30)] = c("Cmp_percentage","TD_percentage","Int_percentage","First_downs","Yards_per_attempt","Adjusted_yards_per_attempt","Yards_per_completion","Yards_per_game","Sck_yards","Sck_percentage","Net_yards_per_attempt","Adjusted_net_yards_per_attempt","Fourth_quarter_comebacks")

## Remove the rank column from the data frame
mod1_temp_passing_df = temp_passing_df %>%
  filter(Rk != "Rk") %>%
  select(-"Rk")

## Change the values of columns with numbers in them to numerical ones
mod2_temp_passing_df = mod1_temp_passing_df %>%
  mutate_if(~any(str_detect(string = ., pattern = '^-?[0-9]\\d*(\\.\\d+)?$')), 
            ~ as.numeric(.))

## Change the NA values in the columns Fourth_quarter_comebacks and GWD to 0
mod3_temp_passing_df = mod2_temp_passing_df %>%
  mutate(Fourth_quarter_comebacks = replace_na(Fourth_quarter_comebacks, 0)) %>%
  mutate(GWD = replace_na(GWD, 0))

## Remove the rows that have NAs in the columns
mod4_temp_passing_df = mod3_temp_passing_df %>%
  na.omit()

## Capitalize the position names in the position column
mod5_temp_passing_df = mod4_temp_passing_df %>%
  mutate(Pos = case_when(
    Pos == 'qb' ~ 'QB',
    Pos == 'wr' ~ 'WR',
    Pos == 'rb' ~ 'RB',
    TRUE ~ Pos
  ))

## Remove row 45, which has a problematic value for team (2TM)
mod5_temp_passing_df = mod5_temp_passing_df[-c(45), ]

## Change the value of the Pos column for Ben Roethlisberger to be QB 
mod6_temp_passing_df = mod5_temp_passing_df %>%
  mutate(Pos = case_when(
    Player == 'Ben Roethlisberger' ~ 'QB',
    TRUE ~ Pos
  ))

## Create a minimum value for passing attempts to filter the data frame by
passing_df = mod6_temp_passing_df %>%
  filter(Att >= 20)
```

## Step 4
Make a visualization with the data from the data frame
```{r}
## Sacks vs Interceptions 
passing_df %>%
  ggplot(aes(x = Sk, y = Int)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
   labs(x = "Sacks", y = "Interceptions Thrown", title = "The Effect Sacks Have On Interceptions Thrown") +
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'), 
        axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold")) +
  theme(plot.background = element_rect(fill = "#ADD8E6"),
        panel.background = element_rect(fill = "#FFCCCB"))
```

## Summary
From this plot, it is clear that there is a strong positive correlation between the number of sacks taken by a QB compared to the interceptions that they throw. As a QB gets sacked a higher number of times, their number of interceptions also increases. This may be as a result of the QB having less time in the pocket to throw a good pass, which could lead to some risky decision making, leading to a higher number of interceptions.
```{r}

## Completion percentage vs First downs 
## Categorized by Age and Game winning drives
passing_df %>%
  ggplot(aes(x = Cmp_percentage, y = First_downs)) +
  geom_jitter(aes(size = GWD, col = Age)) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(x = 'Completion Percentage', y = 'First Downs', title = 'Completion Percentage Compared to First Downs Gained', subtitle = 'Categorized by Age and Game Winning Drives') +
  guides(size = guide_legend(title = 'Game Winning Drives')) +
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'), 
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold"),
        legend.title = element_text(face = 'bold')) +
  theme(plot.background = element_rect(fill = "#90EE90"),
        panel.background = element_rect(fill = "#FFFF00"),
        legend.background = element_rect(fill = '#90ee90'))
```

## Summary
In this visualization, it is clear that there is also a positive relationship between completion percentage and the number of first downs a QB had in a season. Additionally, QBs who had a high number of first downs also had more game winning drives on the season. Something to notice among the QBs who had a high number of game winning drives was that they tended to be older than other QBs in the NFL, as shown by the number of light blue dots amongst QBs who had a high number of game winning drives.  

```{r}
## Create a new column for passing_df for the plot to work properly
passing_df_col1 = passing_df %>%
  mutate(Career_stage = case_when(
    Age > 20 & Age <= 26 ~ 'Young QB',
    Age > 26 & Age <= 33 ~ 'Prime QB',
    Age > 33 & Age <= 45 ~ 'Veteran QB'
  ))

passing_df_col1 %>%
  ggplot(aes(x = Career_stage, y = Yards_per_attempt)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = 'Career Stage', y = 'Yards Per Attempt', title = 'Career Stage of Quarterbacks Compared to Their Yards Per Attempt') +
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'), 
        axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold")) +
  theme(plot.background = element_rect(fill = "#FFA500"),
        panel.background = element_rect(fill = "#ADD8E6"))
```

## Summary
From this visualization, something surprising that jumps out is the fact that veteran QBs, who are defined as QBs from ages 33 to 45, have the highest median yards per attempt, and the lowest range of yards per attempt. This is unexpected, as older QBs are usually expected to have weaker arms, and as a result, attempt shorter throws. However, this plot shows how that assumption is false. Additionally, young QBs, defined as QBs from the age of 20 to 26, have the lowest median yards per attempt.

## Table of the last visualization
```{r}
## Group the passing data frame with the Career_stage column by Career_stage
passing_df_table = passing_df_col1 %>%
  group_by(Career_stage)

## Add columns for the mean YPA, sd YPA, and sample size of YPA for each stage
by_stage = passing_df_table %>%
  summarise(Mean_YPA = mean(Yards_per_attempt),
            SD_YPA = sd(Yards_per_attempt),
            Size_YPA = length(Yards_per_attempt))

## Create the table using cable and custom column names
career_table = kable(by_stage, col.names = c("Career Stage", "Mean Yards Per Attempt", "Standard Deviation Yards Per Attempt", "Sample Size"))

## Change the styling of the table so that it looks better
career_table %>%
  kable_styling(full_width = FALSE)
```

```{r}
## Add an additional table with a key for the viewer to understand the career stages
stage_key = data.frame(
  Stages = c("Prime QB", "Veteran QB", "Young QB"),
  Definition = c(
    "Quarterbacks who are older than 26 and less than or equal to the age of 33 ",
    "Quarterbacks who are older than 33 and less than or equal to the age of 45 ", 
    "Quarterbacks who are older than 20 and less than or equal to the age of 26 "
  )
)

kable(stage_key) %>%
  kable_styling(full_width = FALSE)
```

## Creating a Function to Create a Data Frame

## Step 1
Create the function
```{r}
## Define the function and its parameters
passing_df_function <- function(year, pass_attempts = 150) {
  ## Create a URL for the passing data
  url <- read_html(paste0('https://www.pro-football-reference.com/years/',year,'/passing.htm'))

  ## Turn the information in the URL into a list
  passing_list = url %>%
    ## You can find the label of the specific table by using the selector gadget to select the table you want to use
    html_nodes('#passing') %>%
    html_table()

  ## Create a temporary data frame to hold the information from the list
  ## This is a temporary data frame because we still have to clean the data once we bring the data   in
  temp_passing_df = passing_list[[1]]
  
  ## For loop to iterate through the years and place NAs in the years where the QBR stat wasn't recorded
  for (i in year) {
    if (i < 2006) {
      temp_passing_df = temp_passing_df %>%
        add_column(QBR = NA,
                   .after = 'Rate')
    } 
  }
  
  ## Renaming the columns 
  colnames(temp_passing_df)[c(11,14,16,17,19,20,21,22,26,27,28,29,30)] = c("Cmp_percentage","TD_percentage","Int_percentage","First_downs","Yards_per_attempt","Adjusted_yards_per_attempt","Yards_per_completion","Yards_per_game","Sck_yards","Sck_percentage","Net_yards_per_attempt","Adjusted_net_yards_per_attempt","Fourth_quarter_comebacks")
      
  
  ## Remove the rank column from the data frame
  mod1_temp_passing_df = temp_passing_df %>%
    filter(Rk != "Rk") %>%
    select(-"Rk")
  
  ## Change the values of columns with numbers in them to numerical ones
  mod2_temp_passing_df = mod1_temp_passing_df %>%
    mutate_if(~any(str_detect(string = ., pattern = '^-?[0-9]\\d*(\\.\\d+)?$')), 
            ~ as.numeric(.))
  
  ## Change the NA values in the columns Fourth_quarter_comebacks and GWD to 0
  mod3_temp_passing_df = mod2_temp_passing_df %>%
    mutate(Fourth_quarter_comebacks = replace_na(Fourth_quarter_comebacks, 0)) %>%
    mutate(GWD = replace_na(GWD, 0))
  
  ## Remove the rows that have NAs in the columns except for QBR
  mod4_temp_passing_df = mod3_temp_passing_df %>%
    drop_na(c("Player","Tm","Age","Pos","G","GS","QBrec","Cmp","Att","Cmp_percentage","Yds","TD","TD_percentage","Int","Int_percentage","First_downs","Lng","Yards_per_attempt","Adjusted_yards_per_attempt","Yards_per_completion","Yards_per_game","Rate","Sk","Sck_yards","Sck_percentage","Net_yards_per_attempt","Adjusted_net_yards_per_attempt","Fourth_quarter_comebacks","GWD"))
  
  ## Capitalize the position names in the position column
  mod5_temp_passing_df = mod4_temp_passing_df %>%
    mutate(Pos = case_when(
    Pos == 'qb' ~ 'QB',
    Pos == 'wr' ~ 'WR',
    Pos == 'rb' ~ 'RB',
    TRUE ~ Pos
    ))
  
  ## Create a minimum value for passing attempts to filter the data frame by
  passing_df = mod5_temp_passing_df %>%
    filter(Att >= pass_attempts) %>%
    ## Adding a year column 
    mutate(Year = year)
  
  
  return(passing_df)
}
```

## Step 2
Use the map_df function to run the function through all the years listed and make one big data frame
```{r}
## Create a data frame using the function
## Using the map_df function to use multiple years 
pass_df = map_df(2003:2021, passing_df_function)
```


## Testing block of code
```{r}
## Define the function and its parameters
  url <- read_html(paste0('https://www.pro-football-reference.com/years/2005/passing.htm'))

  ## Turn the information in the URL into a list
  passing_list = url %>%
    ## You can find the label of the specific table by using the selector gadget to select the table you want to use
    html_nodes('#passing') %>%
    html_table()

  ## Create a temporary data frame to hold the information from the list
  ## This is a temporary data frame because we still have to clean the data once we bring the data   in
  temp_passing_df = passing_list[[1]]
  
  temp_passing_df = temp_passing_df %>%
    add_column(QBR = NA,
                .after = 'Rate')
  
  
  ## Renaming the columns 
  colnames(temp_passing_df)[c(11,14,16,17,19,20,21,22,26,27,28,29,30)] = c("Cmp_percentage","TD_percentage","Int_percentage","First_downs","Yards_per_attempt","Adjusted_yards_per_attempt","Yards_per_completion","Yards_per_game","Sck_yards","Sck_percentage","Net_yards_per_attempt","Adjusted_net_yards_per_attempt","Fourth_quarter_comebacks")
  
## Remove the rank column from the data frame
  mod1_temp_passing_df = temp_passing_df %>%
    filter(Rk != "Rk") %>%
    select(-"Rk")
  
  ## Change the values of columns with numbers in them to numerical ones
  mod2_temp_passing_df = mod1_temp_passing_df %>%
    mutate_if(~any(str_detect(string = ., pattern = '^-?[0-9]\\d*(\\.\\d+)?$')), 
            ~ as.numeric(.))
  
  ## Change the NA values in the columns Fourth_quarter_comebacks and GWD to 0
  mod3_temp_passing_df = mod2_temp_passing_df %>%
    mutate(Fourth_quarter_comebacks = replace_na(Fourth_quarter_comebacks, 0)) %>%
    mutate(GWD = replace_na(GWD, 0))
  
  mod4_temp_passing_df = mod3_temp_passing_df %>%
    drop_na(c("Player","Tm","Age","Pos","G","GS","QBrec","Cmp","Att","Cmp_percentage","Yds","TD","TD_percentage","Int","Int_percentage","First_downs","Lng","Yards_per_attempt","Adjusted_yards_per_attempt","Yards_per_completion","Yards_per_game","Rate","Sk","Sck_yards","Sck_percentage","Net_yards_per_attempt","Adjusted_net_yards_per_attempt","Fourth_quarter_comebacks","GWD"))
```



