---
title: "nfl_sports_reference_tutorial"
author: "Siddharth Venkiteswaran"
date: '2022-07-28'
output: html_document
---

## Step 1
Load in the libraries needed to extract and clean the data
```{r}
library(tidyverse)
library(readxl)
library(htmltab)
library(rvest)
```

## Step 2
Go into pro-football-reference.com and select the table you want to work with, and then import that data into R and turn it into a data frame

In this example, I will be using the data that contains the statistics of NFL Quarterbacks from the 2021 NFL season

Make sure you have the selector gadget Chrome extension to select the table you want to use
```{r}
## Create a URL for the passing data
passing_data_link <- read_html('https://www.pro-football-reference.com/years/2021/passing.htm')

## Turn the information in the URL into a list
passing_data_list = passing_data_link %>%
  ## You can find the label of the specific table by using the selector gadget to select the table you want to use
  html_nodes('#passing') %>%
  html_table()

## Create a temporary data frame to hold the information from the list
## This is a temporary data frame because we still have to clean the data once we bring the data in
temp_passing_df = passing_data_list[[1]]
```

## Step 3
Once the data from the URL is turned into a data frame, it's time to clean the data so that we can use it to make visualizations and perform statistical methods with it
```{r}
## Remove the duplicate names in the temporary data frame
## We cannot clean the data until all of the columns have unique names
colnames(temp_passing_df)[c(11,14,16,17,19,20,21,22,26,27,28,29,30)] = c("Cmp_percentage","TD_percentage","Int_percentage","First_downs","Yards_per_attempt","Adjusted_yards_per_attempt","Yards_per_completion","Yards_per_game","Sck_yards","Sck_percentage","Net_yards_per_attempt","Adjusted_net_yards_per_attempt","Fourth_quarter_comebacks")

## Remove the rank column from the data frame
mod1_temp_passing_df = temp_passing_df %>%
  filter(Rk != "Rk") %>%
  select(-"Rk")

## Change the values of columns with numbers in them to numerical ones
mod2_temp_passing_df = mod1_temp_passing_df %>%
  mutate_if(~any(str_detect(string = ., pattern = '^-?[0-9]\\d*(\\.\\d+)?$')), 
            ~ as.numeric(.))

## Change the NA values in the columns Fourth_quarter_comebacks and GWD to 0
mod3_temp_passing_df = mod2_temp_passing_df %>%
  mutate(Fourth_quarter_comebacks = replace_na(Fourth_quarter_comebacks, 0)) %>%
  mutate(GWD = replace_na(GWD, 0))

## Remove the rows that have NAs in the columns
mod4_temp_passing_df = mod3_temp_passing_df %>%
  na.omit()

## Capitalize the position names in the position column
mod5_temp_passing_df = mod4_temp_passing_df %>%
  mutate(Pos = case_when(
    Pos == 'qb' ~ 'QB',
    Pos == 'wr' ~ 'WR',
    Pos == 'rb' ~ 'RB',
    TRUE ~ Pos
  ))

## Remove row 45, which has a problematic value for team (2TM)
mod5_temp_passing_df = mod5_temp_passing_df[-c(45), ]
```

